---
title: "Project proposal"
author: "Richie Rines"
output: github_document
---

```{r load-packages, message = FALSE}
library(tidyverse)
library(broom)
```

## 1. Introduction
The dataset I plan on using for my project is called “Olympic Swimming History (1912 to 2020)”. I found it on Kaggle and the link to it is https://www.kaggle.com/datasets/datasciencedonut/olympic-swimming-1912-to-2020?resource=download. I picked this dataset because swimming is a decent part of my life so I think this will be fun to analyze. It has 4359 observations and 10 variables. The variables are Location, Year, Distance (in meters), Stroke, Relay?, Gender, Team, Athlete, Results, and Rank. The dataset stores all of the Olympic swimming events ranging from 1912 to 2020. I believe I can make a whole bunch of visualizations with the data, including bar graphs, facets, tables, histograms, etc. Some questions I could answer in my project are what strokes have the fastest times, which countries were the fastest, how male times compared with female times, which countries were disqualified the most, is disqualification associated with one gender more than the other?

## 2. Data
```{r}
data <- read.csv("Olympic_Swimming_Results_1912to2020.csv")
```

```{r}
data %>%
  glimpse()
```

## 3. Data analysis plan

**History of Swimming Olympics**
This dataset covers all of the Olympic games (swimming events) from the ones back in 1912 to the most recent ones in 2020. ***NOTE: The 2024 Olympics will be starting in July 2024, a few months after this report was made. Some information shown here will become dated.*** How have the games evolved over the past century? How many events did there used to be and how many are there now?

```{r, echo=FALSE}
#| label: event_num

#didWin variable created here so it is easier to single out the events. there are multiple swimmers in each event but only one can win, so singling out the winners gives us just one entry for each event
data <- data %>%
  mutate(didWin = if_else(Rank == 1, "yes", "no"))

data %>%
  filter(didWin == "yes", Year == "1912" | Year == "2020") %>%
  group_by(Year) %>%
  count() %>%
  select(Year, n) %>%
  summarise(Events = n)
```
Wow! There were only eight events in the 1912 Olympic games! And in 2020, we had 28 total events. That means that over the past 100 years, 20 new events have been added to the Olympics. I wonder what they were?


**How common is each stroke in the Olympic Games? That is, how often does each stroke appear?**
Swimming consists of four main strokes: Backstroke, Breaststroke, Butterfly, and Freestyle. There are also two seperate events, called medleys, that are combinations of all four strokes. Individual medley (IM) is Buttefly, Backstroke, Breaststroke, and then Freestyle. Medley is Backstroke, Breastroke, Butterfly, and then Freestyle. 

```{r echo=FALSE}
#| label: strk_freq

data %>%
  ggplot(aes(x = Stroke, fill = Gender)) +
  geom_bar() +
  scale_fill_manual(values = c("#F98CE5","#77F3FF")) +
  labs(x = "Stroke", 
       y = "Frequency", 
       title = "How Frequent Each Stroke is in the Olympics", 
       subtitle = "And the genders that swim them")
```

***The answer is clearly Freestyle.***
Just by looking at the graph, we can immediately notice that freestyle skyrockets in comparison to the other five strokes. We can also see that the gender (whether a man or woman is swimming) is split almost right down the middle, and that goes for all the strokes. This is to be expected, as for every event there is for a man, there is also one for a woman. Backstroke, breaststroke, and butterfly are evenly matched in frequency, but why is Freestyle so high up? Let's break it down.

```{r}




```

**Which countries are the most successful at the Olympics?**
Many countries all over the world compete in the swimming events, but they can't all get first place. The following table shows the top five countries in the history of the Olympic games in terms of first placements and how many gold medals they won.


```{r, echo=FALSE}
#| label: top_wins

data %>%
  filter(didWin == "yes") %>%
  group_by(Team) %>%
  count(didWin) %>%
  select(Team, n) %>%
  summarise(Wins = n) %>%
  arrange(desc(Wins)) %>%
  slice(1, 2, 3, 4, 5)
```
***USA dominates the competition in the Olympics!***
Taking a look at the table, we see that USA is in the lead with 240 total wins. Australia is next in the podium with 61 wins. Then comes East Germany in third place with 30 wins. Tailing in fourth and fifth are Hungary and Japan, with 25 and 24 wins respectively. 

**When did USA pick up their wins?**

```{r, echo=FALSE}
#| label: usa_wins

data %>%
  mutate(USAwinlose = if_else(Rank == 1, "Wins", "Losses")) %>%
  filter(Team == "USA") %>%
  ggplot(aes(y = USAwinlose, fill = USAwinlose)) +
  geom_bar(color = "black", size = 0.3) +
  scale_fill_manual(values = c("#FF0505","#74EF00")) +
  facet_wrap(. ~ Year) +
  labs(title = "USA Placements in Swimming Events Over the Years",
       subtitle = "From 1912 to 2020",
       x = "Wins/Losses",
       y = "",
       fill = "")
```

At first glance, it looks like all the USA does is lose! But starting in 1960, the country was racking in at least 10 wins every four years (excluding 1988), and all those gold medals add up. Also, when we look at the same graph but with the second-place country, Australia, we see that USA really does have an absurd amount of wins. 

```{r, echo=FALSE}
#| label: aus_wins

data %>%
  mutate(AUSwinlose = if_else(Rank == 1, "Wins", "Losses")) %>%
  filter(Team == "AUS") %>%
  ggplot(aes(y = AUSwinlose, fill = AUSwinlose)) +
  geom_bar(color = "black", size = 0.3) +
  scale_fill_manual(values = c("#ED294A","#3FE241")) +
  facet_wrap(. ~ Year) +
  labs(title = "Australia (AUS) Placements in Swimming Events Over the Years",
       subtitle = "From 1912 to 2020 (with different colors to further seperate from previous graph)",
       x = "Wins/Losses",
       y = "",
       fill = "")
```

Australia has a lot less green, and a whole lot more red! However, they have been somewhat consistent with their placements, and they appear to be on the rise when we look at their wins from 1996 to 2020 (as long as we ignore 2012). 


**Which athlete has the most collective wins?**












```{r, echo=FALSE, eval = FALSE}
data %>%
  filter(didWin == "yes", Year == "2020") %>%
  count()

data %>%
  filter(didWin == "yes", Year == "1964")

data %>%
  count(didWin)

data %>%
  filter(didWin == "yes") %>%
  group_by(Team) %>%
  count(didWin) %>%
  select(Team, n) %>%
  summarise(Wins = n) %>%
  arrange(desc(Wins))
```


```{r, echo=FALSE, eval = FALSE}
data <- data %>%
  mutate(didWin = if_else(Rank == 1, "yes", "no"))
```

```{r, echo=FALSE, eval = FALSE}
data %>%
  filter(Location == "Tokyo", didWin == "yes", Relay. == 0) %>%
  select(Athlete, Results, Location, Year)

```

```{r, echo=FALSE, eval = FALSE}
data %>%
  count(didWin)

data %>%
  filter(Relay. == 0, Stroke != "Medley") %>%
  select(Team, Stroke, Distance..in.meters., didWin)
  
```





































