---
title: "Project proposal"
author: "Richie Rines"
output: github_document
---

```{r load-packages, message = FALSE}
library(tidyverse)
library(broom)
```

``` {r, echo=FALSE}
#| label: colorize

#this code chunk does not matter too much. i just wanted to be able to change the color of some text if i felt like it. i found this code online here: (https://bookdown.org/yihui/rmarkdown-cookbook/font-color.html)

colorize <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{%s}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'>%s</span>", color,
      x)
  } else x
}
```

## 1. Introduction
The dataset I plan on using for my project is called “Olympic Swimming History (1912 to 2020)”. I found it on Kaggle and the link to it is https://www.kaggle.com/datasets/datasciencedonut/olympic-swimming-1912-to-2020?resource=download. I picked this dataset because swimming is a decent part of my life so I think this will be fun to analyze. It has 4359 observations and 10 variables. The variables are Location, Year, Distance (in meters), Stroke, Relay?, Gender, Team, Athlete, Results, and Rank. The dataset stores all of the Olympic swimming events ranging from 1912 to 2020. I believe I can make a whole bunch of visualizations with the data, including bar graphs, facets, tables, histograms, etc.

## 2. Data
```{r}
data <- read.csv("Olympic_Swimming_Results_1912to2020.csv")
```

```{r}
data %>%
  glimpse()
```

## 3. Data analysis plan

**History of Swimming Olympics**

This dataset covers all of the Olympic games (swimming events) from the ones back in 1912 to the most recent ones in 2020. `r colorize("***NOTE: The 2024 Olympics will be starting in July 2024, a few months after this report was made. Some information shown here will become dated.***", "#707070")` How have the games evolved over the past century? How many events did there used to be and how many are there now?

`r colorize("***NOTE: THIS DATASET DOES NOT INCLUDE RELAYS FROM 2020. SOME GRAPHICS MAY DISPLAY INFORMATION FROM 2016 FOR THIS REASON***", "#FF3636")`

```{r, echo=FALSE}
#| label: event_num

#didWin variable created here so it is easier to single out the events. there are multiple swimmers in each event but only one can win, so singling out the winners gives us just one entry for each event
data <- data %>%
  mutate(didWin = if_else(Rank == 1, "yes", "no"))

# I use 2016 here because the dataset does not display the results of the 2020 relays
data %>%
  filter(didWin == "yes", Year == "1912" | Year == "2016") %>%
  group_by(Year) %>%
  count() %>%
  select(Year, n) %>%
  summarise(Events = n)
```

Wow! There were only eight events in the 1912 Olympic games! And in 2016, we had 31 total events. That means that over the past 100 years, over 20 new events have been added to the Olympics. I wonder what they were?

```{r, echo=FALSE}
data %>%
  filter(didWin == "yes", Year == "1912") %>%
  select(Year, Distance..in.meters., Stroke, Gender)

# I use 2016 here because the dataset does not display the results of the 2020 relays
data %>%
  filter(didWin == "yes", Year == "2016", Gender == "Men") %>%
  select(Year, Distance..in.meters., Stroke, Gender)
```

`r colorize("*Although it does not show it, every race has an event for Men and Women. So while the table only displays 16 events, there are really 32. Half of them were hidden to conserve space on the report*", "#F35D5D")`

Looking at the table, we can see that some new events were created over the years. We see the appearance of the 50m and the 200m, as well as individual medley races and relays (medleys). A big reason as to why there was such a drastic increase in events is because after time, all the events that were originally just for men became open to women as well. 

**How common is each stroke in the Olympic Games? That is, how often does each stroke appear?**

Swimming consists of four main strokes: Backstroke, Breaststroke, Butterfly, and Freestyle. There are also two separate events, called medleys, that are combinations of all four strokes. Individual medley (IM) is Butterfly, Backstroke, Breaststroke, and then Freestyle. Medley is Backstroke, Breaststroke, Butterfly, and then Freestyle. 

```{r echo=FALSE}
#| label: strk_freq

data %>%
  ggplot(aes(x = Stroke, fill = Gender)) +
  geom_bar(color = "black", linewidth = 0.3) +
  scale_fill_manual(values = c("#77F3FF","#F98CE5")) +
  labs(x = "Stroke", 
       y = "Frequency", 
       title = "How Frequent Each Stroke is in the Olympics", 
       subtitle = "And the genders that swim them")
```

***The answer is clearly Freestyle.***
Just by looking at the graph, we can immediately notice that freestyle skyrockets in comparison to the other five strokes. We can also see that the gender (whether a man or woman is swimming) is split almost right down the middle, and that goes for all the strokes. This is to be expected, as for every event there is for a man, there is also one for a woman. Backstroke, breaststroke, and butterfly are evenly matched in frequency, but why is Freestyle so high up? Let's break it down.

```{r, echo=FALSE}
#| label: free_other
  
data %>%
  mutate(TempStroke = if_else(Stroke == "Freestyle", "Freestyle", "Other")) %>%
  filter(didWin == "yes", Year == "2016") %>%
  count(TempStroke)
```

According to the table, there are almost just as many freestyle events as there are other events! No wonder the freestyle bar was so high up. 

**Which countries are the most successful at the Olympics?**

Many countries all over the world compete in the swimming events, but they can't all get first place. The following table shows the top five countries in the history of the Olympic games in terms of first placements and how many gold medals they won.


```{r, echo=FALSE}
#| label: top_wins

data %>%
  filter(didWin == "yes") %>%
  group_by(Team) %>%
  count(didWin) %>%
  select(Team, n) %>%
  summarise(Wins = n) %>%
  arrange(desc(Wins)) %>%
  slice(1, 2, 3, 4, 5)
```

USA dominates the competition in the Olympics!
Taking a look at the table, we see that USA is in the lead with 240 total wins. Australia is next in the podium with 61 wins. Then comes East Germany in third place with 30 wins. Tailing in fourth and fifth are Hungary and Japan, with 25 and 24 wins respectively. 

**When did USA pick up their wins?**

```{r, echo=FALSE}
#| label: usa_wins

data %>%
  mutate(USAwinlose = if_else(Rank == 1, "Wins", "Losses")) %>%
  filter(Team == "USA") %>%
  ggplot(aes(y = USAwinlose, fill = USAwinlose)) +
  geom_bar(color = "black", linewidth = 0.3) +
  scale_fill_manual(values = c("#FF0505","#74EF00")) +
  facet_wrap(. ~ Year) +
  labs(title = "USA Placements in Swimming Events Over the Years",
       subtitle = "From 1912 to 2020",
       x = "Wins/Losses",
       y = "",
       fill = "")
```

`r colorize("***Note: the USA did not participate in the 1980 Olympic games due to the boycott created to protest the Soviet Union's invasion in Afghanistan***", "#12C87B")`

At first glance, it looks like all the USA does is lose! But starting in 1960, the country was racking in at least 10 wins every four years (excluding 1988), and all those gold medals add up. Also, when we look at the same graph but with the second-place country, Australia, we see that USA really does have an absurd amount of wins. 

```{r, echo=FALSE}
#| label: aus_wins

data %>%
  mutate(AUSwinlose = if_else(Rank == 1, "Wins", "Losses")) %>%
  filter(Team == "AUS") %>%
  ggplot(aes(y = AUSwinlose, fill = AUSwinlose)) +
  geom_bar(color = "black", linewidth = 0.3) +
  scale_fill_manual(values = c("#ED294A","#3FE241")) +
  facet_wrap(. ~ Year) +
  labs(title = "Australia (AUS) Placements in Swimming Events Over the Years",
       subtitle = "From 1912 to 2020 (with different colors to further seperate from previous graph)",
       x = "Wins/Losses",
       y = "",
       fill = "")
```

Australia has a lot less green, and a whole lot more red! However, they have been somewhat consistent with their placements, and they appear to be on the rise when we look at their wins from 1996 to 2020 (as long as we ignore 2012). 


**Which athlete has the most collective wins?**

```{r echo=FALSE}
data %>%
  filter(didWin == "yes") %>%
  group_by(Athlete) %>%
  count(didWin) %>%
  select(Athlete, n) %>%
  summarise(Wins = n) %>%
  arrange(desc(Wins)) %>%
  slice(1, 2, 3, 4, 5, 6, 7, 8, 9)
```

`r colorize("*^table displays top five win amounts throughout the olympics; entries 4-9 are tied for 4th place^*", "#0A9A6E")`

Looking at the table, we can see that Michael Phelps (from Team `r colorize("**U**", "red")``r colorize("**S**", "white")``r colorize("**A**", "blue")`) has the most individual gold medals out of all the athletes with a total of 13. Katie Ledecky (also from Team `r colorize("**U**", "red")``r colorize("**S**", "white")``r colorize("**A**", "blue")`) is next in line but she still only has six wins, less than half of what Phelps has. Michael Phelps is truly the king of the swimming portion of the summer olympics!



```{r, echo=FALSE, eval = FALSE}
data %>%
  filter(didWin == "yes", Year == "2020") %>%
  count()

data %>%
  filter(didWin == "yes", Year == "1964")

data %>%
  count(didWin)

data %>%
  filter(didWin == "yes") %>%
  group_by(Team) %>%
  count(didWin) %>%
  select(Team, n) %>%
  summarise(Wins = n) %>%
  arrange(desc(Wins))
```


```{r, echo=FALSE, eval = FALSE}
data <- data %>%
  mutate(didWin = if_else(Rank == 1, "yes", "no"))
```

```{r, echo=FALSE, eval = FALSE}
data %>%
  filter(Location == "Tokyo", didWin == "yes", Relay. == 0) %>%
  select(Athlete, Results, Location, Year)

```

```{r, echo=FALSE, eval = FALSE}
data %>%
  count(didWin)

data %>%
  filter(Relay. == 0, Stroke != "Medley") %>%
  select(Team, Stroke, Distance..in.meters., didWin)
  
```





































